---
title: "MC PS1 R Markdown"
author: "Mackenzie Cuffe, 26338157"
date: "2026-09-24"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE, warning=FALSE}
# Packages needed for this assignment.
# install.packages() only needs to be run once ever on your machine -
# it's commented out here so knitting doesn't try to reinstall every time.

# install.packages("tidyverse")
# install.packages("GGally")

library(tidyverse)
library(GGally)
```

# Problem 1

## Education

A school counselor was curious about the average IQ of the students in her
school and took a random sample of 25 students' IQ scores. The following is
the data set below. We want to find a 90% confidence interval for the average
student IQ in the school.

```{r p1-data}
y <- c(105, 69, 86, 100, 82, 111, 104, 110, 87, 108, 87, 90, 94, 113, 112,
       98, 80, 97, 95, 111, 114, 89, 95, 126, 98)
```

To find the confidence interval you must find the mean, variance, standard
deviation, standard error, and margin of error.

### Step 1: Find the mean

```{r p1-mean}
sum_y <- sum(y)
print(sum_y)

length_y <- length(y)
print(length_y)

# mean = sum_y / length_y
mean_y <- sum_y / length_y
print(mean_y)
```

Alternatively you can find the mean by going `mean(y)`, as seen below:

```{r p1-mean-builtin}
mean(y)
```

### Step 2: Find the standard deviation

First you need to get the sum of every individual data value minus the mean,squared. To find this we can use a loop to go through each data value,
subtract the mean, then square the result.

```{r p1-sd-loop}
data_minus_mean <- numeric(length(y))
for (i in 1:length(y)) {
  data_minus_mean[i] <- y[i] - mean(y)
}
print(data_minus_mean)

squared_data <- data_minus_mean ^ 2
print(squared_data)
round(squared_data, digits = 2)

sum_SD <- sum(squared_data)
print(sum_SD)
```

Now that we have the sum of every individual data value subtracted from the
mean and squared, we need `sum_SD` over the number of data values in this
sample minus 1, i.e. `length(y) - 1`.

```{r p1-sd}
variance_y <- sum_SD / (length_y - 1)
print(variance_y)

standard_deviation <- sqrt(variance_y)
print(standard_deviation)
```

Alternatively you can also use `sd(x)`:

```{r p1-sd-builtin}
sd_formula <- sd(y)
print(sd_formula)
```

### Step 3: Find the confidence interval

So now that we have our mean and our standard deviation, we can find the
confidence interval for the average student IQ. The first step is finding the Standard Error (SE), which is just the SD over the square root of our sample number.

```{r p1-se}
SE <- standard_deviation / sqrt(length_y)
print(SE)
```

The next step requires finding the critical value t*, which uses the sample
number minus 1 (24 degrees of freedom), and a two-tailed significance level
of alpha = 0.10, giving 0.05 in each tail.

```{r p1-critical-value}
critical_value <- abs(qt(p = 0.05, df = 24))
print(critical_value)
```

Next we need to find the margin of error, which is the critical value
multiplied by the Standard Error.

```{r p1-margin-error}
margin_error <- critical_value * SE
print(margin_error)
```

The last step in finding the confidence interval is to subtract and add the
margin of error from the mean.

```{r p1-ci}
lower_limit <- mean(y) - margin_error
upper_limit <- mean(y) + margin_error

print(lower_limit)
print(upper_limit)
```

You can state with 90% confidence that the true average IQ of the entire
student population falls somewhere between 93.96 and 102.92.

## Hypothesis test

Next, the school counselor was curious whether the average student IQ in her
school is higher than the average IQ score (100) among all schools in the
country. Using the same sample, conduct the appropriate hypothesis test with
alpha = 0.05.

- Null Hypothesis (H0): the school's average IQ is less than or equal to the national average.
- Alternative Hypothesis (H1): the school's average IQ is higher than the
  national average.

The first step is to find the t-value using the same standard error and mean from the confidence interval above: the sample mean minus the hypothetical mean, divided by the standard error.

```{r p1-t-test}
national_average <- 100
t_test <- (mean(y) - national_average) / SE
print(t_test)
```

This measures how far the sample data landed from the target (the
hypothetical/national average). The data landed 1.012 standard errors below
the target.

Since the alternative hypothesis (H1) is that the school's average IQ (98.44) is higher than the national average (100), we look at the upper tail.

```{r p1-verdict}
# Logical true/false check: did our t-test beat the critical value?
if (t_test <= critical_value) {
  print("Hypothesis Outcome: FAIL TO REJECT the null hypothesis.")
  print("Why: the calculated t-test did not exceed the critical value.")
} else {
  print("Hypothesis Outcome: REJECT the null hypothesis.")
  print("Why: the calculated t-test exceeded the critical value.")
}
```

# Problem 2

## Importing and exploring the data

```{r p2-import}
expenditure <- read.table(
  "https://raw.githubusercontent.com/ASDS-TCD/StatsI_2026/main/datasets/expenditure.txt",
  header = TRUE
)
```
I want to see what I am working with for the table - this allows me to see the basic stats.

Very much like the summary but returns only the first 6 rows within the data, so I can see Y, X1, X2, X3 as their actual data points; this also lets me see State and Region, otherwise without it the state would be treated as a data point.
```{r p2-explore}
summary(expenditure)
head(expenditure)
```

## Relationships among Y, X1, X2, and X3

I'm only pulling out the numeric data points, as State and Region don't
belong in a scatterplot matrix.

```{r p2-cor}
expenditure_data1 <- expenditure[, c("Y", "X1", "X2", "X3")]
round(cor(expenditure_data1), 2)
```

```{r p2-ggpairs, fig.width=8, fig.height=6}
plot_ggpairs <- ggpairs(expenditure_data1)
plot_ggpairs

ggsave("ggpairs_matrix.pdf", plot = plot_ggpairs, width = 8, height = 6)
```

*Found inspiration to chart the plot/graph like this via the ggplot2/GGally
documentation on rdocumentation.org.*

The relationships among Y, X1, X2, and X3 are displayed in the graph with the correlations. From the graph, you can see that X1, X2, and X3 each have a positive relationship with per capita shelter/housing expenditures (Y). The correlation between Y and X1, or personal income, has the strongest
relationship with shelter expenditure data (r = 0.53), followed by
X3/urbanization (r = 0.46) and financial insecurity (X2, r = 0.45). Each of
these datasets represents states with higher income, more financially
insecure residents, and/or more urban populations spending more per capita on housing assistance. Yet, none of these variables represents all of the
variation in spending - the data in each panel resembles a moderate upward
trend in spending.

However, among the variables, the strongest correlation in the matrix is
between income (X1) and urbanization (X3), with a correlation of r = 0.60,
meaning that states with more residents living in urban areas have a higher
income per capita. As the relationship between X1 and X3 is stronger than
either variable's relationship with Y, it may be worth investigating
multicollinearity, as a lot of the information about X1 and X3 may overlap.
From the graph, the lower correlation scores between variables such as income and financial insecurity (X1-X2, r = 0.21), and urbanization and financial insecurity (X2-X3, r = 0.22), suggest that financial insecurity may be an umbrella term for complexities not represented well within the data, such as debt or higher cost of living, rather than immediately associating financial insecurity with lower income.

Overall, the matrix gives a plausible understanding of the states. For
example, wealthier, more urban states may have higher tax bases that allocate more resources towards establishing social service infrastructure.

## Housing expenditure by region

```{r p2-region-boxplot, fig.width=7, fig.height=5}
expenditure_labeled <- expenditure
expenditure_labeled$Region <- factor(
  expenditure$Region,
  labels = c("Northeast", "North Central", "South", "West")
)

plot_boxplot <- ggplot(expenditure_labeled, aes(x = Region, y = Y)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(title = "Housing Expenditure by Region",
       x = "Region",
       y = "Per capita housing expenditure")

plot_boxplot

ggsave("boxplot_region.pdf", plot = plot_boxplot, width = 7, height = 5)
```

```{r p2-region-means}
aggregate(Y ~ Region, data = expenditure_labeled, FUN = mean)
```

According to the boxplots above, the region with the highest per capita
expenditure on housing assistance is the West (Region 4). Its mean sits at
88.3, whereas the next highest is the North Central region (Region 2) with an average of 83.9.

## Y vs X1

```{r p2-y-x1, fig.width=7, fig.height=5}
plot_y_x1 <- ggplot(expenditure, aes(x = X1, y = Y)) +
  geom_point() +
  labs(title = "Housing Expenditure vs Personal Income",
       x = "Per capita personal income (X1)",
       y = "Per capita housing expenditure (Y)")

plot_y_x1

ggsave("scatter_Y_X1.pdf", plot = plot_y_x1, width = 7, height = 5)
```

The graph between Y and X1 shows a positive relationship. This was already
established in the correlation matrix above, which showed a moderately
positive correlation (r = 0.53). Therefore, as income rises across various
states and regions, housing expenditure tends to rise as well, hence the
positive correlation. However, the relationship isn't a perfectly tight one, as the points scatter loosely upward rather than sitting on a straight line.This means income explains some, but not all, of the variation in housing spending across states.

## Y vs X1 by region

```{r p2-y-x1-region, fig.width=7, fig.height=5}
plot_y_x1_region <- ggplot(expenditure_labeled,
                            aes(x = X1, y = Y, color = Region, shape = Region)) +
  geom_point(size = 3) +
  labs(title = "Housing Expenditure vs Personal Income by Region",
       x = "Per capita personal income (X1)",
       y = "Per capita housing expenditure (Y)",
       color = "Region",
       shape = "Region")

plot_y_x1_region

ggsave("scatter_Y_X1_region.pdf", plot = plot_y_x1_region, width = 7, height = 5)
```
